---
title: "SURV 727 HW 5"
author: "Mathew Hill"
date: "2024-11-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You may work in pairs or individually for this assignment. Make sure you join a group in Canvas if you are working in pairs. Turn in this assignment as an HTML or PDF file to ELMS. Make sure to include the R Markdown or Quarto file that was used to generate it. Include the GitHub link for the repository containing these files.
```{r}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
library(tidycensus)
library(censusapi)
library(maps)
library(ggplot2)
library(dplyr)
library(factoextra)
library(ggdendro)
```

# Exploring ACS Data

## In this notebook, we use the Census API to gather data from the American Community Survey (ACS).
```{r}
cs_key <- trimws(read_file("C:/Users/mathe/OneDrive/Documents/census-key.txt"))
#Reading API key from .txt file

census_api_key(cs_key) #establishing it has API key
```

```{r}
acs_il_c <- getCensus(name = "acs/acs5", 
                      vintage = 2016, 
                      vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"), 
                      region = "county:*", regionin = "state:17", key = cs_key) %>% 
  rename(pop = B01003_001E, hh_income = B19013_001E, income = B19301_001E)
head(acs_il_c)
```

## Pull map data for Illinois into a data frame.
```{r}
il_map <- map_data("county", region = "illinois")
head(il_map)
```

## Join the ACS data with the map data. Note that il_map has a column subregion which includes county names. We need a corresponding variable in the ACS data to join both data sets. This needs some transformations, among which the function tolower() might be useful. Call the joined data acs_map.
```{r}
acs_il_c <- acs_il_c %>%
  mutate(subregion = tolower(gsub(" County, Illinois", "", NAME)))

acs_map <- il_map %>%
  left_join(acs_il_c, by = "subregion")

head(acs_map)
```

## After you do this, plot a map of Illinois with Counties colored by per capita income.
```{r}
ggplot(acs_map) + 
  geom_polygon(aes(x = long, y = lat, group = group, fill = income))
```

# Hierarchical Clustering

## We want to find clusters of counties that are similar in their population, average household income and per capita income. First, clean the data so that you have the appropriate variables to use for clustering. Next, create the distance matrix of the cleaned data. This distance matrix can be used to cluster counties, e.g. using the ward method. Plot the dendrogram to find a reasonable number of clusters. Draw boxes around the clusters of your cluster solution. Visualize the county clusters on a map. For this task, create a new acs_map object that now also includes cluster membership as a new column. This column should be called cluster.
```{r}
clustering_data <- acs_il_c %>%
  select(pop, hh_income, income) %>%
  na.omit()  

distance_matrix <- dist(clustering_data, method = "euclidean")

hc <- hclust(distance_matrix, method = "ward.D2")

plot(hc, main = "Dendrogram of Illinois Counties", sub = "", cex = 0.6)

acs_il_c$cluster <- cutree(hc, k = 3)  

acs_map <- il_map %>%
  left_join(acs_il_c, by = "subregion")

ggplot(acs_map, aes(long, lat, group = group, fill = factor(cluster))) +
  geom_polygon(color = "white") +
    coord_fixed(1) +
  labs(title = "Illinois Counties Clustered by Income and Population") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

# Census Tracts

## For the next section we need ACS data on a census tract level. We use the same variables as before.
```{r}
acs_il_t <- getCensus(name = "acs/acs5", 
                      vintage = 2016, 
                      vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"), 
                      region = "tract:*", 
                      regionin = "state:17", 
                      key = cs_key) %>% 
  mutate_all(~ifelse(. == -666666666, NA, .)) %>%  
  rename(pop = B01003_001E, 
         hh_income = B19013_001E, 
         income = B19301_001E)
head(acs_il_t)
```

# k-Means

## As before, clean our data for clustering census tracts based on population, average household income and per capita income.
```{r}
acs_il_t <- acs_il_t %>%
  mutate(
    subregion = tolower(gsub("Census Tract [0-9.]+, | County, Illinois", "", NAME))
  ) %>%
  na.omit()  

head(acs_il_t)

```

## Since we want to use K Means in this section, we start by determining the optimal number of K that results in Clusters with low within but high between variation. Plot within cluster sums of squares for a range of K (e.g. up to 20).
```{r}
clustering_data <- acs_il_t %>%
  select(pop, hh_income, income) %>%
  scale()  

wss <- sapply(1:20, function(k) {
  kmeans(clustering_data, centers = k, nstart = 10)$tot.withinss
})

plot(1:20, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters (K)",
     ylab = "Total Sum of Squares Within Each Cluster",
     main = "Elbow Method")
abline(v = which.min(diff(wss) > 0), col = "red", lty = 2)
```
K = 3 will be optimal




