---
title: "SURV 727 HW 5"
author: "Mathew Hill"
date: "2024-11-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You may work in pairs or individually for this assignment. Make sure you join a group in Canvas if you are working in pairs. Turn in this assignment as an HTML or PDF file to ELMS. Make sure to include the R Markdown or Quarto file that was used to generate it. Include the GitHub link for the repository containing these files.
```{r}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
library(tidycensus)
library(censusapi)
library(maps)
library(ggplot2)
library(dplyr)
library(factoextra)
library(ggdendro)
```

# Exploring ACS Data

## In this notebook, we use the Census API to gather data from the American Community Survey (ACS).
```{r}
cs_key <- trimws(read_file("C:/Users/mathe/OneDrive/Documents/census-key.txt"))
#Reading API key from .txt file

census_api_key(cs_key) #establishing it has API key
```

```{r}
acs_il_c <- getCensus(name = "acs/acs5", 
                      vintage = 2016, 
                      vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"), 
                      region = "county:*", regionin = "state:17", key = cs_key) %>% 
  rename(pop = B01003_001E, hh_income = B19013_001E, income = B19301_001E)
head(acs_il_c)
```


